Protected Areas Practical

### 2024-11-27 tutorial/MSc_PA_practical_answers.Rmd
knitted docs
made corrections

Protected areas are a cornerstone of conservation. In fact, the world has committed to protecting 30% of the planet by 2030. But how do we make that protection meaningful? This practical will use structured decision-making to protect land in Madagascar based on your species distribution maps from your previous practical. The software we will use to do this is `prioritizr`.

@@

Replace with the path where you downloaded `MSc_Protected_Areas`. This means that your code will run from this directory. Make sure you use `/` or `\\` in your path name, otherwise you will get an error.

@@

- no cost: cost is equal across all planning units, which means the prioritisation will only rely on the biodiversity data. However, a protected area plan that doesn't account for cost (and therefore people) is not likely to be very helpful.
- directly estimated: we can do an economic evaluation about economic losses of converting land to protected land. This can be through the cost of purchasing land, costing the staff needed to patrol, the revenue from tourism, the loss of opportunities (loss of mining land, grazing land, agricultural land, and even homes).
- indirectly estimated: we can estimate what land is most costly. For example, the human footprint index is often used to see where people are using land. We then make the assumption that areas that people use are most costly and create the greatest conflict, and should be avoided as a place to put a protected area.

Here, it's an exercise, so we use the Human footprint layer, which is downloaded from this paper:
https://www.nature.com/articles/s41597-022-01284-8#Sec12

@@

## How many planning units were protected per solution?

@@

## How many were not protected 30%
```{r}
print(mean(p1_target_coverage$met) * 100)
```
# Exercise 2

How many species have 70% of their range protected?

@@

## Modify the problem to add PAs in

@@

```{r}
# need to reset budget because it was modified in exercise 2
budget <- terra::global(cost, "sum", na.rm = TRUE)[[1]] * 0.3

p2 <-  p1 %>%
  add_locked_in_constraints(PAs) # this is the only new line

@@

# calculate number of selected planning units by solution
eval_n_summary(p2, s2)

# calculate total cost of solution
eval_cost_summary(p2, s2)

@@

```{r}
# need to reset budget because it was modified in exercise 2
budget <- terra::global(cost, "sum", na.rm = TRUE)[[1]] * 0.3

# create new problem with boundary penalties added to it
p3 <-
  p2 %>%
  add_boundary_penalties(penalty = 0.003, edge_factor = 0.5)

@@

# check percentage of the features that have their target met given the solution
print(mean(p3_target_coverage$met) * 100)

@@

# print scores
print(rc)

@@

# need to reset budget because it was modified in exercise 1
budget <- terra::global(cost, "sum", na.rm = TRUE)[[1]] * 0.3

@@

# plot the solution
par(mfrow=c(2,2))# split into 3 plots side by side
plot(s1, main = "Baseline", axes = FALSE)
plot(s2, main = "With PAs", axes = FALSE)
plot(s4, main = "With PAs + Avoid cities", axes = FALSE)
plot(s3, main = "With PAs + Boundary layer", axes = FALSE)
par(mfrow=c(1,1))# go back to only 1 plot

@@

# check percentage of the features that have their target met given the solution
print(mean(p4_target_coverage$met) * 100)
```

>>>>

Protected areas are a cornerstone of conservation. In fact, the world has committed to protecting 30% of the planet by 2030. But how do we make that protection meaningful? This practical will use structured decision-making to protect land in Madagascar based on your species distribution maps from your previous practical. The software we will use to do this is `prioritizr`. This is the main paper about using prioritizr for systematic conservation planning: https://conbio.onlinelibrary.wiley.com/doi/full/10.1111/cobi.14376

@@

**Ignore this section unless you run into problems accessing files later (e.g. species.tif)**

If you have any problems it could be related to your path. Check what folder you are working from by using

```r
getwd()
```
If the folder path you are working from is different from the path where you downloaded `MSc_Protected_Areas` then you can change it using the setwd() code below. Make sure you use your own path (not mine) and use `/` or `\\` in your path name, otherwise you will get an error.

@@

- no cost: cost is equal across all planning units, which means the prioritisation will only rely on the biodiversity data. However, a protected area plan that doesn't account for cost (and therefore people) is not likely to be very helpful.
- directly estimated: we can evaluate the economic losses of converting land to  protected land. This can be through the cost of purchasing land, the staff needed to patrol, the revenue from tourism, and the loss of opportunities (loss of mining land, grazing land, agricultural land, and even homes).
- indirectly estimated: we can estimate what land is most costly. For example, the human footprint index is often used to see where people use land. We then make the assumption that areas that people use are most costly and create the greatest conflict, and should be avoided as a place to put a protected area.

Here, it's an exercise, so we use the Human footprint layer as a proxy for an indirect cost estimate. The layer was downloaded from this paper: https://www.nature.com/articles/s41597-022-01284-8#Sec12

@@

## How many planning units were protected in this solution?

@@

## What percentage of species had at least 30% of their range protected?
```{r}
print(mean(p1_target_coverage$met) * 100)
```
# Exercise 2

What would be the solution if we require a minimum of 70% of the species range to be protected?

@@

## Modify the problem to add PAs

@@

```{r}
# create new problem where protected areas are locked in
p2 <-  p1 %>%
  add_locked_in_constraints(PAs) # this is the only new line

@@

# calculate the number of selected planning units in the solution
eval_n_summary(p2, s2)

# calculate the total cost of the solution
eval_cost_summary(p2, s2)
# Note that the scenario is more expensive - why do you think that is?
# It is likely the first scenario has no constraints and protects the cheapest sites first 
# that meet the biodiversity target. However, some of these sites are already protected in scenario 2, 
# so the algorithm has to consider sites that are not already protected, which are more expensive
# in order to meet the conservation scenario.
# The more we protect, the more expensive it gets because we use the cheaper sites at the
# beginning, and add more and more costly sites as we go on

@@

```{r}
# create a new problem with boundary penalties added to it
p3 <-
  p2 %>%
  add_boundary_penalties(penalty = 0.003, edge_factor = 0.5)

@@

# check the percentage of the features that have their target met given the solution 
print(mean(p3_target_coverage$met) * 100)

@@

@@

@@

# plot the solution
par(mfrow=c(2,2))# split into 4 plots side by side
plot(s1, main = "Baseline", axes = FALSE)
plot(s2, main = "With PAs", axes = FALSE)
plot(s4, main = "With PAs + Avoid cities", axes = FALSE)
plot(s3, main = "With PAs + Boundary layer", axes = FALSE)
par(mfrow=c(1,1))# go back to only 1 plot

@@

# check the percentage of the features that have their target met given the solution
print(mean(p4_target_coverage$met) * 100)
```

### 2024-11-27 tutorial/MSc_PA_practical_exercise.Rmd
knitted docs
made corrections

Replace with the path where you downloaded `MSc_Protected_Areas`. This means that your code will run from this directory. Make sure you use `/` or `\\` in your path name, otherwise you will get an error.

```
setwd("C:/Downloads/MSc_Protected_Areas-main/MSc_Protected_Areas-main/")
```

@@

```{r}
# need to reset budget because it was modified in exercise 2
budget <- terra::global(cost, "sum", na.rm = TRUE)[[1]] * 0.3

# create new problem with boundary penalties added to it
p3 <-
  p2 %>%
  add_boundary_penalties(penalty = 0.003, edge_factor = 0.5)

@@

# need to reset budget because it was modified in exercise 1
budget <- terra::global(cost, "sum", na.rm = TRUE)[[1]] * 0.3

# create new problem with boundary penalties added to it
p4 <-
  p2 %>%
  # add add_locked_out_constraints(locked_out) here

@@

# plot the solution
par(mfrow=c(2,2))# split into 3 plots side by side
plot(s1, main = "Baseline", axes = FALSE)
plot(s2, main = "With PAs", axes = FALSE)
plot(s4, main = "With PAs + Avoid cities", axes = FALSE)
plot(s3, main = "With PAs + Boundary layer", axes = FALSE)
par(mfrow=c(1,1))# go back to only 1 plot
```

>>>>

**Ignore this section unless you run into problems accessing files later (e.g. species.tif)**

If you have any problems it could be related to your path. Check what folder you are working from by using

```r
getwd()
```
If the folder path you are working from is different from the path where you downloaded `MSc_Protected_Areas` then you can change it using the setwd() code below. Make sure you use your own path (not mine) and use `/` or `\\` in your path name, otherwise you will get an error.

```r
setwd("C:/Downloads/MSc_Protected_Areas-main/MSc_Protected_Areas-main/")
```

@@

```{r}

# create a new problem with boundary penalties added to it
p3 <-
  p2 %>%
  add_boundary_penalties(penalty = 0.003, edge_factor = 0.5)

@@

# create new problem with boundary penalties added to it
p4 <-
  p2 %>%
  # add your locked out constraint here: add_locked_out_constraints(locked_out)

@@

# plot the solution
par(mfrow=c(2,2))# split into 4 plots side by side
plot(s1, main = "Baseline", axes = FALSE)
plot(s2, main = "With PAs", axes = FALSE)
plot(s4, main = "With PAs + Avoid cities", axes = FALSE)
plot(s3, main = "With PAs + Boundary layer", axes = FALSE)
par(mfrow=c(1,1))# go back to only 1 plot
```
